---
title: "Stage 1: Spatial Diagnostics & Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MET Diagnostics & Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Before running complex Factor Analytic models on Multi-Environment Trial (MET) data, it is crucial to inspect the dataset for:
1.  **Connectivity**: Do years or trials share enough germplasm to allow for genetic linkage?
2.  **Spatial Integrity**: Are there missing plots or irregular layouts that need padding?
3.  **Temporal Trends**: Is there a genetic trend or yield drag over time?

The `UtilityFunctions` package provides a suite of Base R visualization tools to perform these checks efficiently.

```{r setup}
library(UtilityFunctions)
```

# 1. Setup Data

We will generate a synthetic MET dataset to demonstrate these features.

```{r data_gen}
set.seed(123)
# Create a dataset with 5 years, 3 locs/year
years <- 2018:2022
data <- data.frame()

for (y in years) {
  for (l in 1:3) {
    # Approx 50 genotypes per trial
    # Some overlap between years (connectivity)
    genos <- paste0("G", sample(1:100, 50))

    trial_df <- data.frame(
      Year = y,
      Trial = paste0("Loc", l, "_", y),
      Genotype = genos,
      Row = rep(1:5, each = 10),
      Column = rep(1:10, 5),
      Yield_bu = rnorm(50, mean = 60 + (y - 2018) * 2, sd = 10) # Genetic gain
    )

    # Introduce some missing spatial data in one trial
    if (y == 2019 && l == 1) {
      trial_df <- trial_df[-c(5, 15, 25), ] # Remove 3 plots
    }

    data <- rbind(data, trial_df)
  }
}

# Convert Yield to t/ha (60 lbs/bu standard)
data$Yield <- data$Yield_bu * 0.06725
head(data)
```

# 2. Connectivity Analysis

Connectivity is vital for multi-year analysis. If two years share no genotypes, they cannot be analyzed in the same single-stage model effectively.

We can check for trial connectivity (detecting disconnected groups/islands) using the `check_met_connectivity` function.

```{r connectivity_check}
conn_res <- check_met_connectivity(data, trial_col = "Year", gen_col = "Genotype")
print(conn_res)
```

# 3. Spatial Integrity

### Visualizing a Trial Map

You can inspect the layout of any specific trial using `plot_spatial` on a filtered subset of the data.

```{r map_raw}
# A complete trial
plot_spatial(data[data$Trial == "Loc1_2018", ],
  row = "Row", col = "Column", fill = "Yield"
)
```

In `Loc1_2019`, we artificially removed plots. Let's visualize it to see the layout.

```{r map_gap}
# A trial with gaps
plot_spatial(data[data$Trial == "Loc1_2019", ],
  row = "Row", col = "Column", fill = "Yield"
)
```

# 4. Temporal Trends

Finally, we check for yield trends over time. The `plot_trend` function provides a ggplot2-based boxplot view showing the spread of data per year and the linear regression trend of the means.

```{r trends}
plot_trend(data, mode = "phenotypic", x = "Year", y = "Yield")
```


